---
title: "GPS22 Project: Spatial Zoning Analysis"
author: "Christopher Janssen"
date: "`r lubridate::today()`"
format:
  html:
    self-contained: true
editor_options:
  chunk_output_type: console
editor:
  markdown:
    wrap: 72
---

# Spatial Zoning Analysis

This notebook performs spatial analysis of GPS clusters against zoning data to classify location types and identify recovery-relevant areas for addiction research.

## Setup

```{r setup}
source(here::here("scripts/r/setup.R"))
source(here::here("scripts/r/database.R"))
```

```{r}
con <- connect_gps_db()
```

## Data Exploration

Let's start by understanding what we have available:

```{r}
# Check existing GPS data
clusters_available <- dbGetQuery(con, "SELECT COUNT(*) as n FROM gps_clusters")[1,1]
geocoding_available <- dbGetQuery(con, "SELECT COUNT(*) as n FROM reverse_geocode_results")[1,1]
```

**Available GPS Data:**
- GPS clusters: `{r} clusters_available`
- Geocoded addresses: `{r} geocoding_available`

```{r}
# Check what tables currently exist
all_tables <- dbListTables(con)
zoning_tables <- all_tables[grepl("zoning|zone", all_tables, ignore.case = TRUE)]
```

**Existing tables:** `{r} paste(all_tables, collapse = ", ")`  
**Zoning-related tables:** `{r} if(length(zoning_tables) > 0) paste(zoning_tables, collapse = ", ") else "None found"`

## Load and Examine Zoning File

```{r}
# Load and clean zoning data in one pipeline
zoning_path <- file.path(path_gps, "zoning/madison_zoning_districts.geojson")

if (!file.exists(zoning_path)) {
  stop("Zoning file not found at: ", zoning_path)
}

madison_zoning <- st_read(zoning_path, quiet = TRUE) |>
  st_make_valid() |>
  # Ensure WGS84 CRS
  (\(x) if (st_crs(x)$epsg != 4326) st_transform(x, 4326) else x)()
```

Found `{r} nrow(madison_zoning)` polygons with `{r} ncol(madison_zoning)` columns: `{r} paste(names(madison_zoning), collapse = ", ")`

```{r}
# Show unique zoning codes
madison_zoning |> 
  filter(!is.na(ZONING_CODE), ZONING_CODE != "") |>
  distinct(ZONING_CODE) |>
  arrange(ZONING_CODE) |>
  knitr::kable(caption = "All Unique Zoning Codes Found")
```

```{r}
# Sample of the data structure
madison_zoning |> 
  select(OBJECTID, ZONING_CODE, ShapeSTArea) |>
  head(10) |>
  st_drop_geometry() |>
  knitr::kable(caption = "Sample Zoning Data")
```

## Design Zoning Categories

Based on the actual zoning codes found, let's create meaningful categories:

```{r}
# Define zoning classification based on actual Madison codes
zoning_categories <- tibble(
  zone_code = c(
    # Residential
    "SR-C1", "SR-C2", "SR-C3", "SR-V1", "SR-V2", "TR-C1", "TR-C2", "TR-C3", "TR-C4", "TR-P", "TR-R", "TR-U1", "TR-U2", "TR-V1", "TR-V2", "DR1", "DR2",
    # Mixed-Use  
    "LMX", "NMX", "RMX", "UMX",
    # Commercial
    "TSS", "CC", "CC-T",
    # Downtown/Urban
    "DC", "UOR", 
    # Employment/Office
    "TE", "SE", "SEC", "EC",
    # Industrial
    "IL", "IG",
    # Special/Institutional
    "A", "AP", "CI", "CN", "PD", "PMHP", "PR", "MC", "THV", "TEMP A", "Temp A"
  ),
  zone_category = c(
    # Residential (17 codes)
    rep("Residential", 17),
    # Mixed-Use (4 codes)  
    rep("Mixed-Use", 4),
    # Commercial (3 codes)
    rep("Commercial", 3),
    # Downtown (2 codes)
    rep("Downtown", 2),
    # Employment (4 codes)
    rep("Employment", 4),
    # Industrial (2 codes)
    rep("Industrial", 2),
    # Special (11 codes)
    rep("Special", 11)
  )
)
```

**Zoning Category Summary:**
```{r}
#| echo: false
category_summary <- zoning_categories |>
  count(zone_category, name = "n_codes") |>
  arrange(desc(n_codes))

knitr::kable(category_summary, caption = "Zoning Categories by Number of Codes")
```

## Prepare Data for Database

```{r}
# Check for unmatched zoning codes
unmatched_check <- madison_zoning |>
  filter(!is.na(ZONING_CODE), ZONING_CODE != "") |>
  left_join(zoning_categories, by = c("ZONING_CODE" = "zone_code")) |>
  filter(is.na(zone_category)) |>
  distinct(ZONING_CODE)

if(nrow(unmatched_check) > 0) {
  cat("Unmatched zoning codes:", paste(unmatched_check$ZONING_CODE, collapse = ", "), "\n")
} else {
  cat("✓ All zoning codes successfully categorized\n")
}

# Only assign to variable since we'll use it multiple times in loading
zoning_for_db <- madison_zoning |>
  filter(!is.na(ZONING_CODE), ZONING_CODE != "") |>
  left_join(zoning_categories, by = c("ZONING_CODE" = "zone_code")) |>
  select(
    objectid = OBJECTID,
    zone_code = ZONING_CODE, 
    zone_category,
    area_sqm = ShapeSTArea,
    geometry
  ) |>
  mutate(zone_category = coalesce(zone_category, "Other"))
```

## Load Data into Existing Table

```{r}
# Check if table already has data
existing_zones <- dbGetQuery(con, "SELECT COUNT(*) as n FROM madison_zoning_districts")[1,1]

# Force reload if we have the wrong number (should be ~2470, not ~43)
force_reload <- existing_zones < 100

if (existing_zones == 0 || force_reload) {
  # Clear existing data first
  dbExecute(con, "DELETE FROM madison_zoning_districts;")
  
  # Convert individual polygons to WKT (keeping them separate!)
  zoning_wkt <- zoning_for_db |>
    # Ensure MultiPolygon format for each individual polygon
    mutate(geometry = st_cast(geometry, "MULTIPOLYGON")) |>
    # Convert to WKT
    mutate(geometry_wkt = st_as_text(geometry)) |>
    st_drop_geometry()
  
  # Insert each individual polygon separately
  insertion_success <- 0
  safe_insert <- safely(~ dbExecute(con, "
    INSERT INTO madison_zoning_districts (objectid, zone_code, zone_category, area_sqm, geom)
    VALUES ($1, $2, $3, $4, ST_SetSRID(ST_GeomFromText($5), 4326));
  ", list(.x$objectid, .x$zone_code, .x$zone_category, .x$area_sqm, .x$geometry_wkt)))
  
  for (i in 1:nrow(zoning_wkt)) {
    row <- zoning_wkt[i, ]
    result <- safe_insert(row)
    
    if (is.null(result$error)) {
      insertion_success <- insertion_success + 1
    } else {
      warning("Failed to insert objectid ", row$objectid, " (", row$zone_code, "): ", result$error$message)
    }
  }
} else {
  insertion_success <- existing_zones
}
```

## Verify Database Table

```{r}
# Show final zoning distribution in database
dbGetQuery(con, "
  SELECT 
    zone_category,
    COUNT(*) as n_districts,
    ROUND(SUM(area_sqm) / 1000000, 1) as total_area_km2
  FROM madison_zoning_districts
  GROUP BY zone_category
  ORDER BY total_area_km2 DESC
") |>
  knitr::kable(caption = "Madison Zoning Districts by Category and Area")
```

## Database Connection Cleanup

```{r}
disconnect_gps_db(con)
```

## Next Steps

Now that we have clean zoning data in the database, we can:

1. **Perform spatial joins** between GPS clusters and zoning districts
2. **Analyze location patterns** by zoning category  
3. **Create recovery-relevant classifications** based on zoning + address data
4. **Generate participant risk profiles** based on location exposure