---
title: "Import OSM Historical Data (2017-2018)"
author: "Christopher Janssen"
date: "`r lubridate::today()`"
format:
  html:
    self-contained: true
editor_options:
  chunk_output_type: console
editor:
  markdown:
    wrap: 72
---

# OSM Historical Data Import

## Overview

GPS2 participant data was collected in 2017-2018. To accurately reflect
the POI landscape during that period, we source OpenStreetMap data from
Geofabrik Wisconsin PBF snapshots rather than a modern Nominatim export.

**Import approach:** `osm2pgsql` with a Flex config (`database/osm/gps2.lua`)

The Lua config serves as both the import driver and schema documentation:
it specifies exactly which OSM tags map to which columns. Using `--create`
makes each import idempotent — tables are dropped and rebuilt from scratch.

**POI classes captured:** `amenity`, `shop`, `leisure`, `tourism`,
`healthcare`, `craft`, `office`, `historic`, `emergency`, `natural`,
`man_made`, `military`

**Data source:** Geofabrik Wisconsin historical extract
(`https://download.geofabrik.de/north-america/us/wisconsin.html`)
— download the `.osm.pbf` for the relevant 2017-2018 date.

**Prerequisites:**

1.  `database/01b-public_data-schema.qmd` has been run (tables exist)
2.  `osm2pgsql >= 1.6` is installed (flex output support required)

## Run Import

Execute the import script from the repository root:

``` bash
# Wisconsin-only PBF (recommended):
./scripts/bash/import-osm.sh /path/to/wisconsin-2018.osm.pbf

# Midwest extract — filter to Wisconsin bbox first:
./scripts/bash/import-osm.sh /path/to/midwest-2018.osm.pbf --filter-wisconsin
```

The script runs `osm2pgsql --create --slim --drop` which fully rebuilds
the tables. After completion, re-run `views/poi-visitable-view.qmd` and
`database/99-create-indexes.qmd`.

## Database Connection

```{r}
source(here::here("scripts/r/setup.R"))
source(here::here("scripts/r/connection.R"))

con <- connect_to_db()
dbGetQuery(con, "SELECT current_database(), current_user")
```

## Verify Load

### Row counts

```{r}
dbGetQuery(con, "
SELECT
    'osm_poi'     AS table_name,
    COUNT(*)      AS row_count
FROM public_data.osm_poi
UNION ALL
SELECT
    'osm_landuse' AS table_name,
    COUNT(*)      AS row_count
FROM public_data.osm_landuse
ORDER BY table_name;
") |> knitr::kable()
```

### Geographic bounding box

The extent should fall within Wisconsin
(lon: -92.9 to -86.2, lat: 42.5 to 47.3).

```{r}
dbGetQuery(con, "
SELECT
    ST_XMin(extent) AS lon_min,
    ST_XMax(extent) AS lon_max,
    ST_YMin(extent) AS lat_min,
    ST_YMax(extent) AS lat_max
FROM (
    SELECT ST_Extent(geom) AS extent
    FROM public_data.osm_poi
) sub;
") |> knitr::kable(digits = 3)
```

### Sample records

```{r}
dbGetQuery(con, "
SELECT osm_id, osm_type, class, type, name, city, lon, lat
FROM public_data.osm_poi
ORDER BY random()
LIMIT 10;
") |> knitr::kable()
```

## Summary Statistics

### POI counts by class

```{r}
dbGetQuery(con, "
SELECT
    class,
    COUNT(*) AS n_poi
FROM public_data.osm_poi
GROUP BY class
ORDER BY n_poi DESC;
") |> knitr::kable()
```

### Top POI types within each class

```{r}
dbGetQuery(con, "
SELECT
    class,
    type,
    COUNT(*) AS n
FROM public_data.osm_poi
GROUP BY class, type
ORDER BY class, n DESC;
") |> knitr::kable()
```

### Landuse counts by type

```{r}
dbGetQuery(con, "
SELECT
    type,
    COUNT(*)          AS n_polygons,
    ROUND(SUM(area_sqkm)::numeric, 2) AS total_area_sqkm
FROM public_data.osm_landuse
GROUP BY type
ORDER BY n_polygons DESC;
") |> knitr::kable()
```

### Visitable POI materialized view

Confirm `osm_poi_visitable` refreshed successfully after the import:

```{r}
dbGetQuery(con, "
SELECT
    'osm_poi'           AS source,
    COUNT(*)            AS count
FROM public_data.osm_poi
UNION ALL
SELECT
    'osm_poi_visitable' AS source,
    COUNT(*)            AS count
FROM public_data.osm_poi_visitable
ORDER BY source;
") |> knitr::kable()
```

```{r}
disconnect_db(con)
```
