---
title: "Enrich GPS Data with Alcohol POI Proximity & Density"
author: "Christopher Janssen"
date: "`r lubridate::today()`"
format:
  html:
    self-contained: true
editor_options:
  chunk_output_type: console
editor:
  markdown:
    wrap: 72
---

# POI Proximity & Density Features

This notebook computes spatial exposure features for extreme-risk
alcohol POI at the GPS point level.

**Two feature types:**

-   **Proximity** — distance (meters) to the nearest POI in each risk
    category, unbounded. Every GPS point gets a value regardless of how
    far away the nearest outlet is.

-   **Density** — count of risk POI within 100m, 250m, and 500m of each
    GPS point.

**Categories:**

-   **on_premise** — bars, pubs, nightclubs, biergartens, breweries
    (consumption venues)
-   **off_premise** — liquor stores, wine shops (retail alcohol sales)

**Input:** `gps_enriched_categorized.csv`

**Output:** `gps_enriched_poi_features.csv`

## Setup

```{r include=FALSE}
source(here::here("scripts/r/setup.R"))
source(here::here("scripts/r/connection.R"))

con <- connect_to_db()
```

## Load GPS Data

```{r}
gps_path <- str_c(path_shared, "/gps_enriched_categorized.csv")

gps_data <- read_csv(gps_path, show_col_types = FALSE)

nrow(gps_data)
```

```{r}
glimpse(gps_data)
```

## Load GPS Points into Temporary Database Table

```{r}
gps_spatial <- gps_data |>
  mutate(row_id = row_number()) |>
  select(row_id, lat, lon)
```

```{r}
dbWriteTable(con, "temp_gps_points", gps_spatial, temporary = TRUE, overwrite = TRUE, row.names = FALSE)
```

```{r}
dbExecute(con, "
ALTER TABLE temp_gps_points
ADD COLUMN geom GEOMETRY(Point, 4326);
")

dbExecute(con, "
UPDATE temp_gps_points
SET geom = ST_SetSRID(ST_MakePoint(lon, lat), 4326);
")
```

```{r}
dbExecute(con, "
CREATE INDEX idx_temp_gps_geom ON temp_gps_points USING GIST(geom);
")

dbExecute(con, "
CREATE INDEX idx_temp_gps_row_id ON temp_gps_points(row_id);
")

dbExecute(con, "ANALYZE temp_gps_points;")
```

```{r}
dbGetQuery(con, "
SELECT COUNT(*) as total_points, COUNT(geom) as with_geometry
FROM temp_gps_points;
")
```

## Verify POI Availability

```{r}
dbGetQuery(con, "
SELECT
  CASE
    WHEN (class = 'amenity' AND type IN ('bar', 'pub', 'nightclub', 'biergarten'))
      OR (class = 'craft' AND type = 'brewery') THEN 'on_premise'
    WHEN class = 'shop' AND type IN ('alcohol', 'wine') THEN 'off_premise'
  END AS category,
  COUNT(*) as n_poi
FROM public_data.osm_poi
WHERE (class = 'amenity' AND type IN ('bar', 'pub', 'nightclub', 'biergarten'))
   OR (class = 'craft' AND type = 'brewery')
   OR (class = 'shop' AND type IN ('alcohol', 'wine'))
GROUP BY category
ORDER BY category;
")
```

## Compute Proximity Features

For each GPS point, find the distance to the nearest on-premise and
off-premise outlet. Unbounded — distance is returned regardless of how
far away the nearest outlet is.

**SQL Pattern:**

-   `CROSS JOIN LATERAL ... ORDER BY geom <-> g.geom LIMIT 1` — uses
    GIST index for fast nearest-neighbor lookup
-   `ST_Distance(...::geography)` — accurate great-circle distance in
    meters

```{r}
proximity_results <- dbGetQuery(con, "
WITH on_premise_poi AS (
  SELECT geom
  FROM public_data.osm_poi
  WHERE (class = 'amenity' AND type IN ('bar', 'pub', 'nightclub', 'biergarten'))
     OR (class = 'craft' AND type = 'brewery')
),
off_premise_poi AS (
  SELECT geom
  FROM public_data.osm_poi
  WHERE class = 'shop' AND type IN ('alcohol', 'wine')
),
nearest_on AS (
  SELECT DISTINCT ON (g.row_id)
    g.row_id,
    ROUND(ST_Distance(g.geom::geography, p.geom::geography)::numeric, 2)
      AS proximity_on_premise_m
  FROM temp_gps_points g
  CROSS JOIN LATERAL (
    SELECT geom FROM on_premise_poi ORDER BY geom <-> g.geom LIMIT 1
  ) p
  ORDER BY g.row_id
),
nearest_off AS (
  SELECT DISTINCT ON (g.row_id)
    g.row_id,
    ROUND(ST_Distance(g.geom::geography, p.geom::geography)::numeric, 2)
      AS proximity_off_premise_m
  FROM temp_gps_points g
  CROSS JOIN LATERAL (
    SELECT geom FROM off_premise_poi ORDER BY geom <-> g.geom LIMIT 1
  ) p
  ORDER BY g.row_id
)
SELECT n1.row_id, n1.proximity_on_premise_m, n2.proximity_off_premise_m
FROM nearest_on n1
JOIN nearest_off n2 ON n1.row_id = n2.row_id;
")

nrow(proximity_results)
```

```{r}
# Summary: proximity distributions (meters)
proximity_results |>
  select(proximity_on_premise_m, proximity_off_premise_m) |>
  summary()
```

## Compute Density Features

For each GPS point, count on-premise and off-premise outlets within
100m, 250m, and 500m.

**SQL Pattern:**

-   LEFT JOIN on 500m (max radius) pre-filters to relevant POI only
-   `COUNT(CASE WHEN ST_DWithin(...) THEN 1 END)` counts within each
    inner radius

```{r}
density_results <- dbGetQuery(con, "
WITH alcohol_poi AS (
  SELECT
    geom,
    CASE
      WHEN (class = 'amenity' AND type IN ('bar', 'pub', 'nightclub', 'biergarten'))
        OR (class = 'craft' AND type = 'brewery') THEN 'on_premise'
      WHEN class = 'shop' AND type IN ('alcohol', 'wine') THEN 'off_premise'
    END AS category
  FROM public_data.osm_poi
  WHERE (class = 'amenity' AND type IN ('bar', 'pub', 'nightclub', 'biergarten'))
     OR (class = 'craft' AND type = 'brewery')
     OR (class = 'shop' AND type IN ('alcohol', 'wine'))
)
SELECT
  g.row_id,
  COUNT(CASE WHEN ST_DWithin(g.geom::geography, p.geom::geography, 100)
                  AND p.category = 'on_premise' THEN 1 END) AS n_on_premise_100m,
  COUNT(CASE WHEN ST_DWithin(g.geom::geography, p.geom::geography, 250)
                  AND p.category = 'on_premise' THEN 1 END) AS n_on_premise_250m,
  COUNT(CASE WHEN ST_DWithin(g.geom::geography, p.geom::geography, 500)
                  AND p.category = 'on_premise' THEN 1 END) AS n_on_premise_500m,
  COUNT(CASE WHEN ST_DWithin(g.geom::geography, p.geom::geography, 100)
                  AND p.category = 'off_premise' THEN 1 END) AS n_off_premise_100m,
  COUNT(CASE WHEN ST_DWithin(g.geom::geography, p.geom::geography, 250)
                  AND p.category = 'off_premise' THEN 1 END) AS n_off_premise_250m,
  COUNT(CASE WHEN ST_DWithin(g.geom::geography, p.geom::geography, 500)
                  AND p.category = 'off_premise' THEN 1 END) AS n_off_premise_500m
FROM temp_gps_points g
LEFT JOIN alcohol_poi p
  ON ST_DWithin(g.geom::geography, p.geom::geography, 500)
GROUP BY g.row_id
ORDER BY g.row_id;
")

# Cast integer64 to integer for R compatibility
density_results <- density_results |>
  mutate(across(starts_with("n_"), as.integer))

nrow(density_results)
```

```{r}
# Summary: density distributions
density_results |>
  select(starts_with("n_")) |>
  summary()
```

## Join All Features

```{r}
gps_enriched_features <- gps_data |>
  mutate(row_id = row_number()) |>
  left_join(proximity_results, by = "row_id") |>
  left_join(density_results, by = "row_id") |>
  # Replace NA density counts with 0 (no POI in range)
  mutate(across(starts_with("n_"), ~replace_na(.x, 0L))) |>
  select(-row_id)

nrow(gps_enriched_features)
```

## Validation

```{r}
# Proximity: no NAs expected (unbounded — every point has a nearest outlet)
tibble(
  feature = c("proximity_on_premise_m", "proximity_off_premise_m"),
  n_na = c(
    sum(is.na(gps_enriched_features$proximity_on_premise_m)),
    sum(is.na(gps_enriched_features$proximity_off_premise_m))
  ),
  min_m = c(
    min(gps_enriched_features$proximity_on_premise_m, na.rm = TRUE),
    min(gps_enriched_features$proximity_off_premise_m, na.rm = TRUE)
  ),
  median_m = c(
    median(gps_enriched_features$proximity_on_premise_m, na.rm = TRUE),
    median(gps_enriched_features$proximity_off_premise_m, na.rm = TRUE)
  ),
  max_m = c(
    max(gps_enriched_features$proximity_on_premise_m, na.rm = TRUE),
    max(gps_enriched_features$proximity_off_premise_m, na.rm = TRUE)
  )
) |>
  knitr::kable()
```

```{r}
# Density: coverage rates at each radius
tibble(
  feature = c(
    "n_on_premise_100m", "n_on_premise_250m", "n_on_premise_500m",
    "n_off_premise_100m", "n_off_premise_250m", "n_off_premise_500m"
  ),
  pct_nonzero = c(
    mean(gps_enriched_features$n_on_premise_100m > 0) * 100,
    mean(gps_enriched_features$n_on_premise_250m > 0) * 100,
    mean(gps_enriched_features$n_on_premise_500m > 0) * 100,
    mean(gps_enriched_features$n_off_premise_100m > 0) * 100,
    mean(gps_enriched_features$n_off_premise_250m > 0) * 100,
    mean(gps_enriched_features$n_off_premise_500m > 0) * 100
  ) |> round(2),
  mean_count = c(
    mean(gps_enriched_features$n_on_premise_100m),
    mean(gps_enriched_features$n_on_premise_250m),
    mean(gps_enriched_features$n_on_premise_500m),
    mean(gps_enriched_features$n_off_premise_100m),
    mean(gps_enriched_features$n_off_premise_250m),
    mean(gps_enriched_features$n_off_premise_500m)
  ) |> round(3),
  max_count = c(
    max(gps_enriched_features$n_on_premise_100m),
    max(gps_enriched_features$n_on_premise_250m),
    max(gps_enriched_features$n_on_premise_500m),
    max(gps_enriched_features$n_off_premise_100m),
    max(gps_enriched_features$n_off_premise_250m),
    max(gps_enriched_features$n_off_premise_500m)
  )
) |>
  knitr::kable()
```

## Write Output

```{r}
output_path <- str_c(path_shared, "/gps_enriched_poi_features.csv")

write_csv(gps_enriched_features, output_path)
```

```{r}
file.info(output_path)$size / 1024^2  # Size in MB
```

## Cleanup

```{r}
dbExecute(con, "DROP TABLE IF EXISTS temp_gps_points;")

disconnect_db(con)
```
