---
title: "Feature Engineer POI Categories"
author: "Christopher Janssen"
date: "`r lubridate::today()`"
format:
  html:
    self-contained: true
editor_options:
  chunk_output_type: console
editor:
  markdown:
    wrap: 72
---

# POI Category Feature Engineering

This notebook creates theory-driven POI categories from raw OSM class/type data for lapse prediction modeling.

## Setup

```{r}
library(tidyverse)
source(here::here("scripts/r/setup.R"))
source(here::here("scripts/r/connection.R"))
```

## Load Enriched Data

```{r}
gps_enriched <- read_csv(str_c(path_shared, "/gps_enriched_all.csv"),
                         show_col_types = FALSE)

nrow(gps_enriched)
```

## Create POI Categories

Three theoretically motivated categories based on direct visits (within 50m):

1. `alcohol_on_premise` — consumption venues (bars, pubs, nightclubs, breweries)
2. `alcohol_off_premise` — retail alcohol sales (liquor stores, wine shops)
3. `healthcare` — medical and recovery-related facilities

All other POI matches are set to NA. A separate binary flag
`alcohol_venue_any` marks visits to either alcohol category.

```{r}
gps_enriched <- gps_enriched |>
  mutate(
    pass_poi_category = case_when(
      is.na(pass_class) ~ NA_character_,

      # alcohol_on_premise: consumption venues
      (pass_class == "amenity" & pass_type %in% c(
        "bar", "pub", "nightclub", "biergarten", "casino", "stripclub"
      )) |
      (pass_class == "craft" & pass_type %in% c(
        "brewery", "distillery", "winery"
      )) ~ "alcohol_on_premise",

      # alcohol_off_premise: retail alcohol sales
      (pass_class == "shop" & pass_type %in% c(
        "alcohol", "wine"
      )) ~ "alcohol_off_premise",

      # healthcare: medical and recovery-related facilities
      (pass_class == "amenity" & pass_type %in% c(
        "hospital", "clinic", "doctors", "dentist", "pharmacy",
        "social_facility", "veterinary", "nursing_home"
      )) |
      (pass_class == "healthcare") |
      (pass_class == "emergency" & pass_type %in% c(
        "emergency_ward_entrance", "ambulance_station"
      )) ~ "healthcare",

      # everything else: not used in modeling
      TRUE ~ NA_character_
    ),

    # binary flag: any direct alcohol venue visit (on or off premise)
    alcohol_venue_any = if_else(
      pass_poi_category %in% c("alcohol_on_premise", "alcohol_off_premise"),
      "yes", "no"
    )
  )
```

## Summary Statistics

### Category counts

```{r}
gps_enriched |>
  count(pass_poi_category, sort = TRUE) |>
  mutate(pct_of_all_gps = round(n / nrow(gps_enriched) * 100, 2)) |>
  knitr::kable()
```

### Alcohol venue any flag

```{r}
gps_enriched |>
  count(alcohol_venue_any) |>
  mutate(pct = round(n / nrow(gps_enriched) * 100, 2)) |>
  knitr::kable()
```

### Types within each category

```{r}
gps_enriched |>
  filter(!is.na(pass_poi_category)) |>
  count(pass_poi_category, pass_class, pass_type, sort = TRUE) |>
  group_by(pass_poi_category) |>
  slice_head(n = 5) |>
  knitr::kable()
```

## Write Updated Data

```{r}
output_path <- str_c(path_shared, "/gps_enriched_categorized.csv")
write_csv(gps_enriched, output_path)
```

```{r}
file.info(output_path)$size / 1024^2  # Size in MB
```
