---
title: "K-means vs DBSCAN Clustering Comparison"
author: "Christopher Janssen"
date: "`r lubridate::today()`"
format:
  html:
    self-contained: true
editor_options:
  chunk_output_type: console
editor:
  markdown:
    wrap: 72
---

# K-means vs DBSCAN Comparison

This notebook compares silhouette-based K-means vs standard DBSCAN for
location clustering using ground truth validation. Ground truth is
filtered to locations with at least one stay within 100m.

## Method Comparison

|   | K-means (PAM) | DBSCAN |
|----|----|----|
| **Requires** | Pre-specify K (or use selection method) | Distance threshold (eps) and minimum points (minPts) |
| **Noise handling** | None—all points assigned to a cluster | Labels low-density points as noise |
| **Cluster shape** | Assumes spherical clusters | Finds arbitrary shapes |
| **Best for** | Known number of groups, compact clusters | Unknown number of clusters, noisy data |
| **GPS use case** | Participant-level clustering (latent profiles) | Location extraction from stay-points |

**Why DBSCAN is standard for GPS:** Human mobility produces dense
clusters (home, work) plus sparse outliers (one-off trips). DBSCAN
discards outliers as noise; K-means forces them into clusters,
distorting centroids or inflating K.

## Setup

```{r}
#| include: false

source(here::here("scripts/r/setup.R"))

library(geosphere)
library(cluster)
library(furrr)
library(leaflet)

plan(multisession, workers = parallel::detectCores() - 1)
theme_set(theme_minimal())
```

## Parameters

```{r}
# Clustering parameters
EPS_METERS <- 50
MIN_STAYS <- 3
K_MAX <- 20

# Ground truth and matching
PROXIMITY_THRESHOLD <- 100    # meters for GT filter and capture
DEDUPE_THRESHOLD <- 50        # meters to merge duplicate GT locations

# Rolling window
WINDOW_INCREMENT <- 7
MIN_SUBJECTS <- 20            # minimum N for reliable window estimates
```

## Load Data

### Stay-Points

Detected stays from raw GPS. Each row is a period where a participant
remained stationary.

```{r}
stays <- read_csv(here::here(path_gps2, "data/stays.csv"), show_col_types = FALSE)

nrow(stays)
```

Add study day relative to each participant's first observation (for
rolling window analysis).

```{r}
stays <- stays |>
  group_by(subid) |>
  mutate(
    study_start = min(start_time),
    study_day = as.numeric(difftime(start_time, study_start, units = "days")) + 1
  ) |>
  ungroup()
```

### Ground Truth Locations

Participant-reported locations (home, work, etc.) collected via survey.

```{r}
ground_truth_raw <- read_csv(
  here::here(path_shared, "locations.csv"),
  show_col_types = FALSE
)

nrow(ground_truth_raw)
```

## Prepare Ground Truth (Validation Data)

Participant-reported locations serve as ground truth. We clean them by
merging nearby reports and filtering to locations with GPS evidence.

### Merge Nearby Reports

Participants sometimes report the same location multiple times. Merge
reports within 50m into a single location using median coordinates.

```{r}
dedupe_locations <- function(df, threshold_m = DEDUPE_THRESHOLD) {
  if (nrow(df) < 2) {
    return(df |> mutate(location_id = 1))
  }

  coords <- cbind(df$lon, df$lat)
  dist_mat <- distm(coords, coords, fun = distHaversine)
  clusters <- dbscan(as.dist(dist_mat), eps = threshold_m, minPts = 1)

  df |>
    mutate(location_id = clusters$cluster) |>
    group_by(location_id) |>
    summarize(
      lat = median(lat),
      lon = median(lon),
      n_reports = n(),
      types = paste(unique(na.omit(type)), collapse = ", "),
      .groups = "drop"
    )
}

ground_truth_deduped <- ground_truth_raw |>
  group_by(subid) |>
  group_modify(~dedupe_locations(.x)) |>
  ungroup()

nrow(ground_truth_deduped)
```

### Filter to Locations with GPS Evidence

Keep only ground truth locations that have at least one stay within
100m. This ensures we're evaluating clustering on locations that *could*
be detected.

```{r}
filter_gt_by_proximity <- function(gt_df, stays_df, threshold_m = PROXIMITY_THRESHOLD) {
  gt_df |>
    rowwise() |>
    mutate(
      has_nearby_stay = {
        subj_stays <- stays_df |> filter(subid == .env$subid)
        if (nrow(subj_stays) == 0) {
          FALSE
        } else {
          gt_coord <- c(lon, lat)
          stay_coords <- cbind(subj_stays$lon, subj_stays$lat)
          min(distHaversine(gt_coord, stay_coords)) <= threshold_m
        }
      }
    ) |>
    ungroup() |>
    filter(has_nearby_stay) |>
    select(-has_nearby_stay)
}

ground_truth <- filter_gt_by_proximity(ground_truth_deduped, stays)
```

Summary of ground truth cleaning:

```{r}
tibble(
  version = c("Raw locations", "After deduplication", "After proximity filter"),
  n_locations = c(
    nrow(ground_truth_raw),
    nrow(ground_truth_deduped),
    nrow(ground_truth)
  ),
  n_subjects = c(
    n_distinct(ground_truth_raw$subid),
    n_distinct(ground_truth_deduped$subid),
    n_distinct(ground_truth$subid)
  )
) |>
  knitr::kable()
```

## Clustering Implementation

Both methods use Haversine distances (geographic, not Euclidean) via a
precomputed distance matrix.

### Run DBSCAN and K-means

**DBSCAN:** Groups points within `eps` meters if at least `minPts`
neighbors exist. Isolated points become noise (cluster = 0).

**K-means (PAM):** Tests K = 2 to 20, selects K with highest silhouette
score. Uses PAM (medoids) instead of standard K-means to work with
distance matrices. Falls back to K=1 if silhouette is weak and data is
spatially compact.

```{r}
run_dbscan <- function(coords, eps = EPS_METERS, min_pts = MIN_STAYS) {
  if (nrow(coords) < 2) {
    return(list(
      cluster = rep(1, nrow(coords)),
      n_clusters = 1,
      n_noise = 0
    ))
  }

  dist_mat <- distm(coords, coords, fun = distHaversine)
  result <- dbscan(as.dist(dist_mat), eps = eps, minPts = min_pts)

  list(
    cluster = result$cluster,
    n_clusters = max(result$cluster),
    n_noise = sum(result$cluster == 0)
  )
}

run_kmeans <- function(coords, k_max = K_MAX) {
  n_points <- nrow(coords)

  if (n_points < 2) {
    return(list(
      cluster = rep(1, n_points),
      n_clusters = 1,
      optimal_k = 1
    ))
  }

  dist_mat <- distm(coords, coords, fun = distHaversine)
  k_max <- min(k_max, n_points - 1)

  if (k_max < 2) {
    return(list(
      cluster = rep(1, n_points),
      n_clusters = 1,
      optimal_k = 1
    ))
  }

  # Silhouette scores for K = 2 to k_max
  k_range <- 2:k_max
  sil_scores <- map_dbl(k_range, function(k) {
    km <- pam(as.dist(dist_mat), k = k, diss = TRUE)
    mean(silhouette(km$clustering, as.dist(dist_mat))[, 3])
  })

  best_k <- k_range[which.max(sil_scores)]
  best_sil <- max(sil_scores)

  # If weak structure and spatially compact, use K=1
  if (best_sil < 0.25) {
    max_dist <- max(dist_mat)
    if (max_dist <= 200) {
      return(list(
        cluster = rep(1, n_points),
        n_clusters = 1,
        optimal_k = 1
      ))
    }
  }

  km <- pam(as.dist(dist_mat), k = best_k, diss = TRUE)

  list(
    cluster = km$clustering,
    n_clusters = best_k,
    optimal_k = best_k
  )
}
```

### Extract Cluster Centers

Compute median lat/lon for each cluster. Excludes DBSCAN noise points
(cluster = 0).

```{r}
get_centroids <- function(stays_df, cluster_col) {
  stays_df |>
    filter(.data[[cluster_col]] > 0) |>
    group_by(.data[[cluster_col]]) |>
    summarize(
      lat = median(lat),
      lon = median(lon),
      total_dwell_hrs = sum(dwell_mins) / 60,
      n_stays = n(),
      n_days = n_distinct(as.Date(start_time)),
      .groups = "drop"
    ) |>
    rename(cluster_id = 1)
}
```

### Calculate Precision and Recall

-   **Recall:** What fraction of ground truth locations have a cluster
    centroid within 100m?
-   **Precision:** What fraction of cluster centroids are within 100m of
    a ground truth location?
-   **F1:** Harmonic mean of precision and recall.

```{r}
calculate_capture <- function(centroids, gt_locations, threshold_m = PROXIMITY_THRESHOLD) {
  if (nrow(centroids) == 0 || nrow(gt_locations) == 0) {
    return(list(
      recall = 0,
      precision = 0,
      f1 = 0,
      n_captured = 0,
      n_valid_centroids = 0,
      n_ground_truth = nrow(gt_locations),
      n_centroids = nrow(centroids)
    ))
  }

  centroid_coords <- cbind(centroids$lon, centroids$lat)
  gt_coords <- cbind(gt_locations$lon, gt_locations$lat)

  # Recall: fraction of GT locations with a nearby centroid
  captured <- 0
  for (i in seq_len(nrow(gt_locations))) {
    distances <- distHaversine(gt_coords[i, ], centroid_coords)
    if (min(distances) <= threshold_m) captured <- captured + 1
  }
  recall <- captured / nrow(gt_locations)

  # Precision: fraction of centroids near a GT location
  valid_centroids <- 0
  for (i in seq_len(nrow(centroids))) {
    distances <- distHaversine(centroid_coords[i, ], gt_coords)
    if (min(distances) <= threshold_m) valid_centroids <- valid_centroids + 1
  }
  precision <- valid_centroids / nrow(centroids)

  # F1 score
  f1 <- if (precision + recall > 0) {
    2 * (precision * recall) / (precision + recall)
  } else {
    0
  }

  list(
    recall = recall,
    precision = precision,
    f1 = f1,
    n_captured = captured,
    n_valid_centroids = valid_centroids,
    n_ground_truth = nrow(gt_locations),
    n_centroids = nrow(centroids)
  )
}
```

## Rolling Window Analysis

Test how clustering performance changes as data accumulates from study
onset. At each 7-day window, we cluster all stays up to that point and
compare against ground truth.

```{r}
analyze_window <- function(sub, window_days, stays, ground_truth) {
  stays_window <- stays |>
    filter(subid == sub, study_day <= window_days)

  n_stays_window <- nrow(stays_window)

  if (n_stays_window < 3) {
    return(NULL)
  }

  gt <- ground_truth |> filter(subid == sub)
  coords <- cbind(stays_window$lon, stays_window$lat)

  # Run clustering

dbscan_result <- run_dbscan(coords)
  kmeans_result <- run_kmeans(coords)

  # Get centroids
  stays_with_dbscan <- stays_window |>
    mutate(cluster_dbscan = dbscan_result$cluster)
  stays_with_kmeans <- stays_window |>
    mutate(cluster_kmeans = kmeans_result$cluster)

  centroids_dbscan <- get_centroids(stays_with_dbscan, "cluster_dbscan")
  centroids_kmeans <- get_centroids(stays_with_kmeans, "cluster_kmeans")

  # Calculate capture scores
  capture_dbscan <- calculate_capture(centroids_dbscan, gt)
  capture_kmeans <- calculate_capture(centroids_kmeans, gt)

  tibble(
    subid = sub,
    window_days = window_days,
    n_stays_window = n_stays_window,
    n_ground_truth = capture_dbscan$n_ground_truth,
    # DBSCAN
    dbscan_clusters = dbscan_result$n_clusters,
    dbscan_noise = dbscan_result$n_noise,
    dbscan_recall = capture_dbscan$recall,
    dbscan_precision = capture_dbscan$precision,
    dbscan_f1 = capture_dbscan$f1,
    # K-means
    kmeans_clusters = kmeans_result$n_clusters,
    kmeans_k = kmeans_result$optimal_k,
    kmeans_recall = capture_kmeans$recall,
    kmeans_precision = capture_kmeans$precision,
    kmeans_f1 = capture_kmeans$f1
  )
}
```

### Create 7-Day Windows

For each subject, create windows at days 7, 14, 21, ... up to their max
study day.

```{r}
subject_durations <- stays |>
  group_by(subid) |>
  summarize(
    max_study_day = max(study_day),
    n_stays = n(),
    .groups = "drop"
  )

windows <- subject_durations |>
  filter(max_study_day >= WINDOW_INCREMENT) |>
  rowwise() |>
  mutate(
    windows = list(seq(WINDOW_INCREMENT, max_study_day, by = WINDOW_INCREMENT))
  ) |>
  unnest(windows) |>
  rename(window_days = windows) |>
  select(subid, window_days)

nrow(windows)
```

### Run Clustering at Each Window (Parallel)

For each subject-window combination, run both clustering methods and
compute metrics against ground truth. Only subjects with ground truth
are included.

```{r}
#| cache: true

analysis_windows <- windows |>
  filter(subid %in% unique(ground_truth$subid))

results <- future_pmap_dfr(
  list(
    sub = analysis_windows$subid,
    window_days = analysis_windows$window_days
  ),
  analyze_window,
  stays = stays,
  ground_truth = ground_truth,
  .options = furrr_options(seed = TRUE),
  .progress = TRUE
)

nrow(results)
```

## Results: Performance Over Time

### Aggregate Metrics by Window

Average precision, recall, and F1 across subjects at each time window.

```{r}
metrics_by_window <- results |>
  group_by(window_days) |>
  summarize(
    n_subjects = n(),
    dbscan_recall_mean = mean(dbscan_recall, na.rm = TRUE),
    dbscan_precision_mean = mean(dbscan_precision, na.rm = TRUE),
    dbscan_f1_mean = mean(dbscan_f1, na.rm = TRUE),
    dbscan_clusters_mean = mean(dbscan_clusters, na.rm = TRUE),
    kmeans_recall_mean = mean(kmeans_recall, na.rm = TRUE),
    kmeans_precision_mean = mean(kmeans_precision, na.rm = TRUE),
    kmeans_f1_mean = mean(kmeans_f1, na.rm = TRUE),
    kmeans_clusters_mean = mean(kmeans_clusters, na.rm = TRUE),
    .groups = "drop"
  )

# Check which windows are excluded
reliable_range <- metrics_by_window |> filter(n_subjects >= MIN_SUBJECTS)
```

```{r}
tibble(
  metric = c("Total windows", "Reliable windows (N >= 20)", "Reliable range (days)"),
  value = c(
    nrow(metrics_by_window),
    nrow(reliable_range),
    paste(min(reliable_range$window_days), "-", max(reliable_range$window_days))
  )
) |>
  knitr::kable()
```

### Recall Over Time

How quickly does each method find ground truth locations? Dashed line =
80% threshold.

```{r}
metrics_by_window |>
  filter(n_subjects >= MIN_SUBJECTS) |>
  pivot_longer(
    cols = c(dbscan_recall_mean, kmeans_recall_mean),
    names_to = "method",
    values_to = "recall"
  ) |>
  mutate(method = if_else(method == "dbscan_recall_mean", "DBSCAN", "K-means")) |>
  ggplot(aes(x = window_days, y = recall, color = method)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_hline(yintercept = 0.8, linetype = "dashed", alpha = 0.5) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_color_manual(values = c("DBSCAN" = "coral", "K-means" = "steelblue")) +
  labs(
    title = "Recall Over Time",
    subtitle = str_glue("Fraction of ground truth locations captured (N >= {MIN_SUBJECTS})"),
    x = "Days from Study Onset",
    y = "Recall",
    color = "Method"
  ) +
  theme(legend.position = "bottom")
```

### Precision Over Time

What fraction of detected clusters correspond to real locations? Low
precision = spurious clusters.

```{r}
metrics_by_window |>
  filter(n_subjects >= MIN_SUBJECTS) |>
  pivot_longer(
    cols = c(dbscan_precision_mean, kmeans_precision_mean),
    names_to = "method",
    values_to = "precision"
  ) |>
  mutate(method = if_else(method == "dbscan_precision_mean", "DBSCAN", "K-means")) |>
  ggplot(aes(x = window_days, y = precision, color = method)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_color_manual(values = c("DBSCAN" = "coral", "K-means" = "steelblue")) +
  labs(
    title = "Precision Over Time",
    subtitle = str_glue("Fraction of clusters matching ground truth (N >= {MIN_SUBJECTS})"),
    x = "Days from Study Onset",
    y = "Precision",
    color = "Method"
  ) +
  theme(legend.position = "bottom")
```

### F1 Score Over Time

Combined measure balancing precision and recall.

```{r}
metrics_by_window |>
  filter(n_subjects >= MIN_SUBJECTS) |>
  pivot_longer(
    cols = c(dbscan_f1_mean, kmeans_f1_mean),
    names_to = "method",
    values_to = "f1"
  ) |>
  mutate(method = if_else(method == "dbscan_f1_mean", "DBSCAN", "K-means")) |>
  ggplot(aes(x = window_days, y = f1, color = method)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_color_manual(values = c("DBSCAN" = "coral", "K-means" = "steelblue")) +
  labs(
    title = "F1 Score Over Time",
    subtitle = str_glue("(N >= {MIN_SUBJECTS})"),
    x = "Days from Study Onset",
    y = "F1 Score",
    color = "Method"
  ) +
  theme(legend.position = "bottom")
```

### Convergence: Days to 80% Recall

How many days until each method captures 80% of ground truth locations?
NA = never reached 80%.

```{r}
days_to_80 <- results |>
  group_by(subid) |>
  summarize(
    dbscan_days = min(window_days[dbscan_recall >= 0.8], na.rm = TRUE),
    kmeans_days = min(window_days[kmeans_recall >= 0.8], na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    dbscan_days = if_else(is.infinite(dbscan_days), NA_real_, dbscan_days),
    kmeans_days = if_else(is.infinite(kmeans_days), NA_real_, kmeans_days)
  )

tibble(
  method = c("DBSCAN", "K-means"),
  subjects_reaching_80 = c(
    sum(!is.na(days_to_80$dbscan_days)),
    sum(!is.na(days_to_80$kmeans_days))
  ),
  mean_days = c(
    round(mean(days_to_80$dbscan_days, na.rm = TRUE), 1),
    round(mean(days_to_80$kmeans_days, na.rm = TRUE), 1)
  ),
  median_days = c(
    round(median(days_to_80$dbscan_days, na.rm = TRUE), 1),
    round(median(days_to_80$kmeans_days, na.rm = TRUE), 1)
  )
) |>
  knitr::kable(col.names = c("Method", "Subjects Reaching 80%", "Mean Days", "Median Days"))
```

### Full Study Comparison

Performance using all available data for each subject (their final
window).

```{r}
final_results <- results |>
  group_by(subid) |>
  filter(window_days == max(window_days)) |>
  ungroup()

tibble(
  method = c("DBSCAN", "K-means"),
  mean_f1 = c(
    paste0(round(mean(final_results$dbscan_f1) * 100, 1), "%"),
    paste0(round(mean(final_results$kmeans_f1) * 100, 1), "%")
  ),
  mean_recall = c(
    paste0(round(mean(final_results$dbscan_recall) * 100, 1), "%"),
    paste0(round(mean(final_results$kmeans_recall) * 100, 1), "%")
  ),
  mean_precision = c(
    paste0(round(mean(final_results$dbscan_precision) * 100, 1), "%"),
    paste0(round(mean(final_results$kmeans_precision) * 100, 1), "%")
  ),
  mean_clusters = c(
    round(mean(final_results$dbscan_clusters), 1),
    round(mean(final_results$kmeans_clusters), 1)
  )
) |>
  knitr::kable(col.names = c("Method", "Mean F1", "Mean Recall", "Mean Precision", "Mean Clusters"))
```

Per-subject comparison. Points above the diagonal = DBSCAN outperformed
K-means for that subject.

```{r}
final_results |>
  ggplot(aes(x = kmeans_f1, y = dbscan_f1)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
  scale_x_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    title = "F1 Score: DBSCAN vs K-means",
    subtitle = "Points above the line = DBSCAN performed better",
    x = "K-means F1 Score",
    y = "DBSCAN F1 Score"
  )
```

## EDA: Interactive Map

Visualize raw GPS, stays, cluster centroids, and ground truth for a
single subject. Toggle layers to compare methods.

```{r}
# Load raw GPS for visualization
gps_raw <- read_csv(
  here::here(path_shared, "gps.csv"),
  show_col_types = FALSE
) |>
  select(subid, time, lat = lat, lon = lon)
```

```{r}
viz_subject <- function(sub, gps_raw, stays, ground_truth) {
  # Filter data for subject

raw <- gps_raw |> filter(subid == sub)
  subj_stays <- stays |> filter(subid == sub)
  gt <- ground_truth |> filter(subid == sub)

  # Run clustering
  coords <- cbind(subj_stays$lon, subj_stays$lat)
  dbscan_result <- run_dbscan(coords)
  kmeans_result <- run_kmeans(coords)

  # Get centroids
  subj_stays <- subj_stays |>
    mutate(
      cluster_dbscan = dbscan_result$cluster,
      cluster_kmeans = kmeans_result$cluster
    )

  centroids_dbscan <- get_centroids(subj_stays, "cluster_dbscan")
  centroids_kmeans <- get_centroids(subj_stays, "cluster_kmeans")

  # Build map
  leaflet() |>
    addProviderTiles(providers$CartoDB.Positron) |>
    # Raw GPS
    addCircleMarkers(
      data = raw, lng = ~lon, lat = ~lat,
      radius = 2, color = "gray", fillOpacity = 0.3, stroke = FALSE,
      group = "Raw GPS"
    ) |>
    # Stays
    addCircleMarkers(
      data = subj_stays, lng = ~lon, lat = ~lat,
      radius = 4, color = "black", fillOpacity = 0.5, stroke = FALSE,
      group = "Stays"
    ) |>
    # DBSCAN centroids
    addCircleMarkers(
      data = centroids_dbscan, lng = ~lon, lat = ~lat,
      radius = 8, color = "coral", fillColor = "coral", fillOpacity = 0.8,
      weight = 2, group = "DBSCAN"
    ) |>
    # K-means centroids
    addCircleMarkers(
      data = centroids_kmeans, lng = ~lon, lat = ~lat,
      radius = 8, color = "steelblue", fillColor = "steelblue", fillOpacity = 0.8,
      weight = 2, group = "K-means"
    ) |>
    # Ground truth
    addCircleMarkers(
      data = gt, lng = ~lon, lat = ~lat,
      radius = 10, color = "darkgreen", fillColor = "green", fillOpacity = 0.7,
      weight = 3, group = "Ground Truth"
    ) |>
    addCircles(
      data = gt, lng = ~lon, lat = ~lat,
      radius = PROXIMITY_THRESHOLD, color = "green", fillOpacity = 0.1, weight = 1,
      group = "GT Capture Radius"
    ) |>
    addLayersControl(
      overlayGroups = c("Raw GPS", "Stays", "DBSCAN", "K-means", "Ground Truth", "GT Capture Radius"),
      options = layersControlOptions(collapsed = FALSE)
    ) |>
    hideGroup("Raw GPS")
}
```

Change the subject ID to explore different participants:

```{r}
viz_subject(10, gps_raw, stays, ground_truth)
```
