---
title: "Verify Adaptive Sampling Assumption"
author: "Christopher Janssen"
date: "`r lubridate::today()`"
format:
  html:
    self-contained: true
editor_options:
  chunk_output_type: console
editor:
  markdown:
    wrap: 72
---

# Verify Adaptive Sampling Hypothesis

This notebook empirically tests whether FollowMe and Moves apps use adaptive sampling (reduced ping frequency when stationary).

TLDR: If adaptive sampling is used, we expect:

-   Longer time gaps between GPS points during stationary periods
-   Shorter time gaps during movement
-   This pattern should be evident in the data

## Setup

```{r}
#| include: false

source(here::here("scripts/r/setup.R"))

library(geosphere)

theme_set(theme_bw())
```

## Load Raw GPS Data with App Source

```{r}
data <- read_csv(here::here(path_shared, "/gps.csv"), show_col_types = FALSE)

nrow(data)
```

## Identify/Check App Sources

```{r}
data |>
  count(app_source, name = "n_points") |>
  mutate(pct = round(n_points / sum(n_points) * 100, 1)) |>
  knitr::kable()
```

## Calculate Movement Metrics

```{r}
# calculate time gaps and distance between consecutive points
data_movement <- data |>
  arrange(subid, time) |>
  group_by(subid) |>
  mutate(
    # time gap from previous point (seconds)
    time_gap_secs = as.numeric(difftime(time, lag(time), units = "secs")),

    # distance from previous point (meters, Haversine)
    dist_from_prev = if_else(
      row_number() == 1,
      NA_real_,
      distHaversine(cbind(lag(lon), lag(lat)), cbind(lon, lat))
    ),

    # classify as stationary or moving
    # stationary: within 50m of previous point
    is_stationary = !is.na(dist_from_prev) & dist_from_prev <= 50,

    # get app_source from current point
    app = app_source
  ) |>
  ungroup()

# check
data_movement |>
  select(subid, time, app_source, time_gap_secs, dist_from_prev, is_stationary) |>
  slice_head(n = 10)
```

## Exclude first point per subject (no lag) and extreme gaps (>1 hour)

```{r}
data_filtered <- data_movement |>
  filter(!is.na(time_gap_secs),
         time_gap_secs > 0,
         time_gap_secs < 3600)  # less than 1 hour

nrow(data_filtered)
```

## Check Impact of 1-Hour Exclusion

### What Are We Excluding?

```{r}
# analyze gaps that were excluded (>= 1 hour)
excluded_gaps <- data_movement |>
  filter(!is.na(time_gap_secs), time_gap_secs >= 3600) |>
  group_by(app_source, is_stationary) |>
  summarize(
    n_excluded = n(),
    median_excluded_hrs = median(time_gap_secs) / 3600,
    max_excluded_hrs = max(time_gap_secs) / 3600,
    .groups = "drop"
  ) |>
  mutate(
    movement = if_else(is_stationary, "Stationary", "Moving")
  ) |>
  select(app_source, movement, everything(), -is_stationary)

excluded_gaps |>
  knitr::kable(digits = 2, caption = "Excluded gaps (>= 1 hour)")
```

### What percentage of data are we excluding?

```{r}
exclusion_rate <- data_movement |>
  filter(!is.na(time_gap_secs), time_gap_secs > 0) |>
  group_by(app_source, is_stationary) |>
  summarize(
    total_gaps = n(),
    excluded_gaps = sum(time_gap_secs >= 3600),
    exclusion_pct = round(100 * sum(time_gap_secs >= 3600) / n(), 2),
    .groups = "drop"
  ) |>
  mutate(
    movement = if_else(is_stationary, "Stationary", "Moving")
  ) |>
  select(app_source, movement, total_gaps, excluded_gaps, exclusion_pct)

exclusion_rate |>
  knitr::kable(caption = "Exclusion rates by app and movement state")
```

### Full Distribution (No Filter)

```{r}
# Compare WITH and WITHOUT the 1-hour filter
full_distribution <- data_movement |>
  filter(!is.na(time_gap_secs), time_gap_secs > 0) |>
  group_by(app_source, is_stationary) |>
  summarize(
    n_all = n(),
    median_all = median(time_gap_secs),
    q95_all = quantile(time_gap_secs, 0.95),
    q99_all = quantile(time_gap_secs, 0.99),
    max_all_hrs = max(time_gap_secs) / 3600,
    .groups = "drop"
  ) |>
  mutate(
    movement = if_else(is_stationary, "Stationary", "Moving")
  ) |>
  select(app_source, movement, everything(), -is_stationary)

full_distribution |>
  knitr::kable(digits = 1, caption = "Full distribution (no filter)")
```

### Calculate ratio using FULL data (no exclusion)

```{r}
full_ratio <- full_distribution |>
  select(app_source, movement, median_all) |>
  pivot_wider(names_from = movement, values_from = median_all) |>
  mutate(
    ratio_full_data = Stationary / Moving
  )

full_ratio |>
  knitr::kable(digits = 2, caption = "Ratio using full data (no 1-hour exclusion)")
```

## Test Adaptive Sampling Hypothesis

### Overall Pattern: Stationary vs. Moving

```{r}
sampling_summary <- data_filtered |>
  group_by(app_source, is_stationary) |>
  summarize(
    n_points = n(),
    min_gap = min(time_gap_secs),
    q25 = quantile(time_gap_secs, 0.25),
    median_gap = median(time_gap_secs),
    q75 = quantile(time_gap_secs, 0.75),
    max_gap = max(time_gap_secs),
    mean_gap = mean(time_gap_secs),
    .groups = "drop"
  ) |>
  mutate(
    movement = if_else(is_stationary, "Stationary", "Moving")
  ) |>
  select(app_source, movement, everything(), -is_stationary)

sampling_summary |>
  knitr::kable(digits = 1)
```

### Interpretation

```{r}
# Calculate ratio of median gaps (stationary / moving)
# If adaptive sampling: ratio > 1 (longer gaps when stationary)
# If fixed sampling: ratio ≈ 1 (same gaps regardless)
interpretation <- sampling_summary |>
  select(app_source, movement, median_gap) |>
  pivot_wider(names_from = movement, values_from = median_gap) |>
  mutate(
    ratio_stationary_to_moving = Stationary / Moving,
    adaptive_sampling = if_else(ratio_stationary_to_moving > 1.5,
                                 "likely adaptive sampling",
                                 "likely fixed-interval sampling")
  )

interpretation |>
  knitr::kable(digits = 2)
```

## Visualizations

### Box Plot Comparison

```{r}
#| echo: false

data_filtered |>
  mutate(
    movement = if_else(is_stationary, "Stationary", "Moving")
  ) |>
  ggplot(aes(x = movement, y = time_gap_secs, fill = movement)) +
  geom_boxplot() +
  facet_wrap(~app_source, ncol = 2) +
  scale_y_log10(breaks = c(10, 30, 60, 120, 300, 600, 1800, 3600),
                labels = c("10s", "30s", "1m", "2m", "5m", "10m", "30m", "1h")) +
  labs(
    title = "Time Gaps: Stationary vs. Moving by App",
    x = "Movement State",
    y = "Time Gap (log scale)",
    fill = "Movement State"
  )
```