---
title: "OSM Data Exploration"
author: "Christopher Janssen"
date: "`r lubridate::today()`"
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 3
    code-fold: show
editor_options:
  chunk_output_type: console
---

# Overview

This notebook explores the Wisconsin OpenStreetMap (OSM) data in the `public_data` schema of the GPS3 geolocation database. The data includes points of interest (POI) and landuse polygons extracted from OpenStreetMap.

## Setup

```{r setup}
#| message: false
#| warning: false

source(here::here("scripts/r/setup.R"))
source(here::here("scripts/r/connection.R"))
```

## Database Connection

Connect to the geolocation database on the shared VM.

```{r}
# Connect to database
con <- connect_to_db()

# Verify connection
dbGetQuery(con, "SELECT current_database(), current_user, version()")
```

## Table Schema

Check actual column structure in the loaded tables.

```{r}
# OSM POI columns
dbGetQuery(con, "
SELECT
    column_name,
    data_type
FROM information_schema.columns
WHERE table_schema = 'public_data' AND table_name = 'osm_poi'
ORDER BY ordinal_position;
")
```

```{r}
# OSM Landuse columns
dbGetQuery(con, "
SELECT
    column_name,
    data_type
FROM information_schema.columns
WHERE table_schema = 'public_data' AND table_name = 'osm_landuse'
ORDER BY ordinal_position;
")
```

# Database Overview

## Table Sizes

```{r}
# Get table sizes for public_data schema
table_sizes <- dbGetQuery(con, "
SELECT
  schemaname,
  tablename,
  pg_size_pretty(pg_total_relation_size(schemaname||'.'||tablename)) AS size
FROM pg_tables
WHERE schemaname = 'public_data'
ORDER BY pg_total_relation_size(schemaname||'.'||tablename) DESC;
")

table_sizes |> knitr::kable()
```

## Overall Statistics

```{r}
# Get counts for OSM tables
stats <- dbGetQuery(con, "
SELECT
  'OSM Points of Interest' as dataset,
  COUNT(*) as count
FROM public_data.osm_poi
UNION ALL
SELECT
  'OSM Landuse Polygons',
  COUNT(*)
FROM public_data.osm_landuse;
")

stats |>
  knitr::kable(format.args = list(big.mark = ","))
```

# POI Classification

The `osm_poi` table contains points of interest organized by `class` and `type`. This creates a hierarchical categorization system.

## POI Class-Level Summary

```{r}
class_summary <- dbGetQuery(con, "
SELECT
  class,
  COUNT(*) as n_places,
  COUNT(DISTINCT type) as n_types,
  ROUND(100.0 * COUNT(*) / SUM(COUNT(*)) OVER (), 2) as pct_of_total
FROM public_data.osm_poi
GROUP BY class
ORDER BY n_places DESC;
")

class_summary |>
  knitr::kable(
    col.names = c("Class", "Places", "Types", "% of Total"),
    format.args = list(big.mark = ",")
  )
```

## Landuse Class Summary

```{r}
landuse_summary <- dbGetQuery(con, "
SELECT
  class,
  COUNT(*) as n_polygons,
  COUNT(DISTINCT type) as n_types,
  ROUND(100.0 * COUNT(*) / SUM(COUNT(*)) OVER (), 2) as pct_of_total
FROM public_data.osm_landuse
GROUP BY class
ORDER BY n_polygons DESC;
")

landuse_summary |>
  knitr::kable(
    col.names = c("Class", "Polygons", "Types", "% of Total"),
    format.args = list(big.mark = ",")
  )
```

# Detailed Type Breakdowns

## Shops

```{r}
shop_types <- dbGetQuery(con, "
SELECT
  type,
  COUNT(*) as n_places
FROM public_data.osm_poi
WHERE class = 'shop'
GROUP BY type
ORDER BY n_places DESC
LIMIT 40;
")

shop_types |>
  knitr::kable(
    col.names = c("Shop Type", "Count"),
    format.args = list(big.mark = ",")
  )
```

## Amenities

```{r}
amenity_types <- dbGetQuery(con, "
SELECT
  type,
  COUNT(*) as n_places
FROM public_data.osm_poi
WHERE class = 'amenity'
GROUP BY type
ORDER BY n_places DESC
LIMIT 40;
")

amenity_types |>
  knitr::kable(
    col.names = c("Amenity Type", "Count"),
    format.args = list(big.mark = ",")
  )
```

## Leisure

```{r}
leisure_types <- dbGetQuery(con, "
SELECT
  type,
  COUNT(*) as n_places
FROM public_data.osm_poi
WHERE class = 'leisure'
GROUP BY type
ORDER BY n_places DESC;
")

leisure_types |>
  knitr::kable(
    col.names = c("Leisure Type", "Count"),
    format.args = list(big.mark = ",")
  )
```

## Tourism

```{r}
tourism_types <- dbGetQuery(con, "
SELECT
  type,
  COUNT(*) as n_places
FROM public_data.osm_poi
WHERE class = 'tourism'
GROUP BY type
ORDER BY n_places DESC;
")

tourism_types |>
  knitr::kable(
    col.names = c("Tourism Type", "Count"),
    format.args = list(big.mark = ",")
  )
```

## Healthcare

```{r}
healthcare_types <- dbGetQuery(con, "
SELECT
  type,
  COUNT(*) as n_places
FROM public_data.osm_poi
WHERE class = 'healthcare'
GROUP BY type
ORDER BY n_places DESC;
")

healthcare_types |>
  knitr::kable(
    col.names = c("Healthcare Type", "Count"),
    format.args = list(big.mark = ",")
  )
```

## Places (Named Locations)

```{r}
place_types <- dbGetQuery(con, "
SELECT
  type,
  COUNT(*) as n_places,
  ROUND(100.0 * COUNT(*) / SUM(COUNT(*)) OVER (), 2) as pct
FROM public_data.osm_poi
WHERE class = 'place'
GROUP BY type
ORDER BY n_places DESC;
")

place_types |>
  knitr::kable(
    col.names = c("Place Type", "Count", "% of Places"),
    format.args = list(big.mark = ",")
  )
```

# Data Structure

The actual columns in your loaded tables are shown in the Table Schema section above. Use that as the reference for available fields.

## Example: Viewing POI Data

```{r}
# Sample amenities from Madison
sample_poi <- dbGetQuery(con, "
SELECT
  osm_id,
  osm_type,
  class,
  type,
  city,
  lat,
  lon
FROM public_data.osm_poi
WHERE class = 'amenity'
  AND type IN ('restaurant', 'bar', 'cafe')
  AND city = 'Madison'
LIMIT 15;
")

sample_poi |> knitr::kable()
```

# Example Queries

## Finding Specific POI Types

### All bars and restaurants in Madison

```{sql eval=FALSE}
SELECT
  osm_id,
  class,
  type,
  city,
  lat,
  lon
FROM public_data.osm_poi
WHERE city = 'Madison'
  AND class = 'amenity'
  AND type IN ('bar', 'restaurant', 'pub', 'cafe');
```

### All healthcare facilities

```{sql eval=FALSE}
SELECT
  osm_id,
  class,
  type,
  city,
  lat,
  lon
FROM public_data.osm_poi
WHERE class IN ('healthcare', 'amenity')
  AND type IN ('hospital', 'clinic', 'doctors', 'dentist', 'pharmacy');
```

### Schools by type

```{sql eval=FALSE}
SELECT
  osm_id,
  CASE
    WHEN type = 'kindergarten' THEN 'Kindergarten'
    WHEN type = 'school' THEN 'K-12 School'
    WHEN type = 'college' THEN 'College'
    WHEN type = 'university' THEN 'University'
    ELSE type
  END as school_type,
  city,
  lat,
  lon
FROM public_data.osm_poi
WHERE class = 'amenity'
  AND type IN ('kindergarten', 'school', 'college', 'university');
```

### Spatial query: POI near a location

```{sql eval=FALSE}
-- Find all amenities within 1km of a specific point
-- (Example: UW-Madison Memorial Union at 43.0766, -89.3984)
SELECT
  osm_id,
  class,
  type,
  city,
  ST_Distance(
    geom::geography,
    ST_SetSRID(ST_MakePoint(-89.3984, 43.0766), 4326)::geography
  ) as distance_meters
FROM public_data.osm_poi
WHERE ST_DWithin(
  geom::geography,
  ST_SetSRID(ST_MakePoint(-89.3984, 43.0766), 4326)::geography,
  1000  -- 1km radius
)
ORDER BY distance_meters
LIMIT 20;
```

## Working with Landuse Polygons

### Find landuse polygons containing a point

```{sql eval=FALSE}
-- Find which landuse polygon(s) contain a specific GPS point
SELECT
  osm_id,
  class,
  type,
  area_sqkm
FROM public_data.osm_landuse
WHERE ST_Contains(
  geom,
  ST_SetSRID(ST_MakePoint(-89.3984, 43.0766), 4326)
)
ORDER BY area_sqkm;
```

### Largest landuse polygons by type

```{sql eval=FALSE}
-- Find the 10 largest parks in Wisconsin
SELECT
  osm_id,
  class,
  type,
  city,
  area_sqkm
FROM public_data.osm_landuse
WHERE class = 'leisure' AND type = 'park'
ORDER BY area_sqkm DESC
LIMIT 10;
```

## Cleanup

```{r}
# Close database connection
disconnect_db(con)
```
